---
title: 'Geog6300: Lab 6'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
library(knitr)
library(car)
library(lmtest)
```

**Overview:**
This lab focuses on regression techniques. You'll be analyzing the association of various physical and climatological characteristics in Australia with observations of several animals recorded on the citizen science app iNaturalist.

###Data and research questions###

Let's import the dataset.

```{r}
lab6_data<-st_read("data/aus_climate_inat.gpkg")
```

The dataset for this lab is a 1 decimal degree hexagon grid that has aggregate statistics for a number of variables:

* ndvi: NDVI/vegetation index values from Landsat data (via Google Earth Engine). These values range from -1 to 1, with higher values indicating more vegetation.
* maxtemp_00/20_med: Median maximum temperature (C) in 2000 or 2020 (data from SILO/Queensland government)
* mintemp_00/20_med: Median minimum temperature (C) in 2020 or 2020 (data from SILO/Queensland government)
* rain_00/20_sum: Total rainfall (mm) in 2000 or 2020 (data from SILO/Queensland government)
* pop_00/20: Total population in 2000 or 2020 (data from NASA's Gridded Population of the World)
* water_00/20_pct: Percentage of land covered by water at some point during the year in 2000 or 2020
* elev_med: Median elevation (meters) (data from the Shuttle Radar Topography Mission/NASA)

There are also observation counts from iNaturalist for several distinctively Australian animal species: the central bearded dragon, the common emu, the red kangaroo, the agile wallaby, the laughing kookaburra, the wombat, the koala, and the platypus.

Our primary research question is how the climatological/physical variables in our dataset are predictive of the NDVI value. We will build models for 2020 as well as the change from 2000 to 2020. The second is referred to as a "first difference" model and can sometimes be more useful for identifying causal mechanisms.

###Part 1: Analysis of 2020 data###

We will start by looking at data for 2020. 

**Question 1** _Create histograms for NDVI, max temp., min temp., rain, and population, and water in 2020 as well as elevation. Based on these graphs, assess the normality of these variables._

```{r}
#Code goes here.
ggplot() + 
  geom_histogram(data = lab6_data, aes(x = ndvi_20_med))
ggplot() + 
  geom_histogram(data = lab6_data, aes(x=maxtemp_20_med))
ggplot() + 
  geom_histogram(data = lab6_data, aes(x = mintemp_20_med))
ggplot() + 
  geom_histogram(data = lab6_data, aes(x=rain_20_sum))
ggplot() + 
  geom_histogram(data = lab6_data, aes(x=pop_20))
ggplot() + 
  geom_histogram(data = lab6_data, aes(x=water_20_pct))
```

First, looking at histogram from the ndvi/vegetation index, the histogram appears to be skewed towards the right which would indicate that the data is not normally distributed. Second, looking at the histogram from the maximum temperature variable, it appears that the histogram is skewed towards the left and therefore is not normally distributed. After this, the minimum temperature histogram is still slightly skewed to the left but less skewed than the maximum temperature histogram. Then, the rain_sum histogram is clearly skewed towards the right and not normally distributed. Lastly, the population histogram appears to be skewed towards the right with a clear spike in counts on the left side of the histogram and small little peaks after the large spike. Therefore, this variable is also not normally distributed based on its histogram!

**Question 2** _Use tmap to map these same variables using Jenks natural breaks as the classification method. For an extra challenge, use `tmap_arrange` to plot all maps in a single figure._

```{r}
library(tmap)

ndvi <- tm_shape(lab6_data) + 
  tm_polygons("ndvi_20_med", style="jenks")
max_temp <- tm_shape(lab6_data) + 
  tm_polygons("maxtemp_20_med", style = "jenks")
min_temp <- tm_shape(lab6_data) + 
  tm_polygons("mintemp_20_med", style = "jenks")
rain_sum <- tm_shape(lab6_data) +
  tm_polygons("rain_20_sum", style = "jenks")
population_2020 <- tm_shape(lab6_data) + 
  tm_polygons("pop_20", style = "jenks")
water <- tm_shape(lab6_data) + 
  tm_polygons("water_20_pct", style = "jenks")

tmap_arrange(ndvi, max_temp, min_temp, rain_sum, population_2020, water)

  
```


**Question 3** _Based on the maps from question 3, summarise major patterns you see in the spatial distribution of these data from any of your variables of interest. How do they appear to be associated with the NDVI variable?_

First looking at the medium maximum temperature over Australia during 2020, the coolest temperatures are located along the Southeastern Coast and Southern Australia at large, with higher median temperatures existing in the northwestern and northern parts of the country. Next, looking at the median minimum temperatures over Australia, the lowest minimum median temperatures exist once again along the Southeastern coast of Australia and Southern Australia at large, with the highest minimum temperatures existing in the Northern and Northwestern parts of the country. The total rainfall variable shows maximums in rain towards the northern parts of the country, the southeastern and eastern portions of the country, and a small little area of increased rainfall in the Southwestern portion of the country. The spatial distribution of the population appears to be centered around cities in the Southwestern coast, the Southeastern Coast, and the Eastern Coast. I belieive that these cities are Perth, Brisbane, Sydney, Melbourne, and Canberra. Lastly, the total percent of water coverage shows higher values in Southwest Australia and directly west of the Eastern coastline of the country. 

Now, taking all of these variables and comparing it to the ndvi variable, which is a measurement of how much vegetation is an area with values closer to one having more vegetation we see that generally cooler maximum and minimum temperatures correspond to higher ndvi values. In addition, higher total rainfall occurs with higher ndvi values (with only an exception in far northern Australia which is more remote jungle like terrain). The areas with the highest population tend to also occur in areas with the highest ndvi indexes. Lastly, it appears that the total amount of water covering land does not as strong an associated with ndvi, however just off the maxmimum water contours we see some increased ndvi contours which could indicate that more fertile land comes from the increased water coverage.  

**Question 4** _Create univariate models for each of the variables listed in question 1, with NDVI in 2020 as the dependent variable. Print a summary of each model. Write a summary of those results that indicates the direction, magnitude, and significance for each model coefficient._

```{r}
model_max_temp <- lm(ndvi_20_med~maxtemp_20_med, data=lab6_data)
model_min_temp <- lm(ndvi_20_med~mintemp_20_med, data=lab6_data)
model_rain <- lm(ndvi_20_med~rain_20_sum, data=lab6_data)
model_population <- lm(ndvi_20_med~pop_20, data=lab6_data)
model_water <- lm(ndvi_20_med~water_20_pct, data=lab6_data)
summary(model_max_temp)
summary(model_min_temp)
summary(model_rain)
summary(model_population)
summary(model_water)

```

The first univariate model explains changes in NDVI with median maxmimum temperatures across Australia. First looking at the intercept, with a maximum median temperature of 0 degrees Celsius we would predict an NDVI index of 0.661238. This value is very significant with a very low p-value, and the direction of this intercept is positive because of the positive NDVI value. Next, looking at the slope, with a one degree celsius increase in maximum temperature we will see the NDVI drop by 0.0130902. This is also a very significant result with a very low p-value, and the direction of this value is negative as the NDVI decreases with every one unit increase in temperature.

The second univariate model explains changes in NDVI with median minimum temperatures across Australia. Looking at the intercept, with a median minimum temperature of zero degrees celsius we will see a NDVI index of 0.4644. This value is very significant because of its very low p-value, and the direction of this value is positive because of its positive magnitude. Next, looking at the slope we see with every one degree celsius increase in median minimum temperatures we see a decrease of 0.012282 in the NDVI index. This is also a very signficant result with a very low p-value, and the direction of thi value is negative because of its negative magnitude

The third univariate model explains changes in NDVI with total rainfall across Australia. Looking at the intercept, with zero total rainfall we would expect a NDVI index of 0.1303. This value is significant with a very low p-value and its direction is positive albeit barely positive. Next, looking at the slope we see with every one mm increase in rainfall we see an increase in the NDVI index of 9.124e-07. This is also a very significant resuls with a very low p-value and a positive direction because of its positive magnitude

The fourth univariate model explains changes in NDVI with total population. Looking at the intercept, we see that with zero population we would expect an NDVI index of 0.2552. This value is significant with a very low p-value and a its direction is positive. Next, looking at the slope, with a every one person increase in population we see an increase in the NDVI index of 1.500e-06. This is also a significant result with a very low p-value and a positve direction. 

The last univariate model explains changes in NDVI with water coverage over land. When there is no water coverage over land, we would expect an NDVI index of 0.268988. This value is significant with a very low p-value and a positive direction. Next, looking at the slope, with every 1% increase in water coverage we would expect the NDVI index to decrease by 0.178263. This value is NOT significant with a p-value of 0.249, and the direction of this is negative!

**Question 5** _Create a multivariate regression model with the variables of interest, choosing EITHER max or min temperature (but not both) You may also choose to leave out any variables that were insignificant in Q4. Use the univa=riate models as your guide. Call the results._

```{r}
#Code goes here
multi_model <- lm(ndvi_20_med ~ maxtemp_20_med + rain_20_sum + pop_20, data=lab6_data)
summary(multi_model)


```

**Question 6** _Summarize the results of the multivariate model. What are the direction, magnitude, and significance of each coefficient? How did it change from the univariate models you created in Q4 (if at all)? What do the R2 and F-statistic values tell you about overall model fit?_

For this multivariate regression model, I decided to remove the water coverage variable because the univariate model showed this variable was not significant. So the variable I examined were maximum temperatures, total rainfall, and population and their effect on the ndvi index. Looking at the intercept, when max_temp, total rainfall, and population are all zero, we would expect an NDVI index of 0.4922. With every one degree celsius increase in maximum temperature we expect the NDVI index to decrease by 0.01179. This is a signficant result because of the very low p-value and has a negative direction because of its negative sign 
With every one mm increase in total rainfall, we would expect the NDVI index to increase by 8.648e-07 which is signficant (becaue of the very low p-value) and has a positive direction (because of the positive sign attached to the slope). Lastly, for every one person increase in population we would expect the NDVI index to increase by 3.024e-07 which is signficant but not as stronly signficant as the other coefficients and has a positive direction.
There was very minimal change in these values in this multivariate model when compared to the results from the individual univariate models. 
The R2 value tells us the percent of variation in the dependent variable that can be explained by our independent variables. In this case the R2 value is 0.6158 which indicates that this data strongly fits the data and is a good model! The F-statistic is very signficant which means that our model improves the data compared to having a model with no independent variables.  

**Question 7** _Use a histogram and a map to assess the normality of residuals and any spatial autocorrelation. Summarise any notable patterns that you see._

```{r}
#Code goes here
lab6_data$residuals<-residuals(multi_model)
ggplot(lab6_data, aes(x=residuals)) + 
  geom_histogram()
tm_shape(lab6_data) + 
  tm_polygons("residuals", style = "jenks")
```

First looking at the histogram of residuals, it appears that are residuals are NOT normally distributed. But instead are skewed towards the left which could be issues with the model. Then, looking at the map, it also appears that the data is not normally distributed with points with very negative residuals scattered around positive residuals. In addition, the residuals are spatially autocorrelated and not random, which indicates there could be issues with the model.

**Question 8** _Assess any issues with multicollinearity or heteroskedastity in this model using the techniques shown in class. Run the appropriate tests and explain what their results show you._

```{r}
#Code goes here
vif(multi_model)
bptest(multi_model)
```

Running VIF to see if we have any issues with multicollinearity, we see that each of the independent variables have values around 1. This means that there is no correlation with the other variable, and means we don't have issues with multicollinearity. 

Next, I ran bptest to see if there are any issues with heteroscedasticity. The null hypothesis is there is no heteroscedasticity and the alternative hypothesis means that there is heteroscedastiticty in the model. Because we have such a small p-value, we reject the null hypothesis and accept the alternative that there is heteroscedasticity in the model. 



**Question 9** _How would you summarise the results of this model in a sentence or two? In addition, looking at the full model and your diagnostics, do you feel this is a model that provides meaningful results? Explain your answer._

Overall I feel like the results from my univariate regression models and my multivariate regression model gives meaningful results. However, there appears to be some issues with the residuals that I think require further examination to see if this model is as effective as I think it could be.


```


**Disclosure of assistance:** _Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?_

I used in-class notes and previous scripts from this class to help complete this lab!

**Lab reflection:** _How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What were the biggest things you learned by completing it?_

I felt like I did pretty well on this lab. I would say this lab was overall easy, but the responses were moderate just because of how much a 


**Challenge question**

#Option 1
Create a first difference model. To do that, subtract the values in 2000 from the values in 2020 for each variable for which that is appropriate. Then create a new model similar to the one you created in question 5, but using these new variables showing the *change in values* over time. Call the results of the model, and interpret the results in the same ways you did above. Also chart and map the residuals to assess model error. Finally, write a short section that summarises what, if anything, this model tells you. 

```{r}

```




#Option 2
The animal data included in this dataset is an example of count data, and usually we would use a Poisson or similar model for that purpose. Let's try it with regular OLS regression though. Create two regression models to assess how the counts of two different animals (say, koalas and emus) are associated with at least three of the environmental/climatological variables given above. Be sure to use the same independent variables in each model. Interpret the results of each model and then explain the importance of any differences in the model coefficients between them, focusing on direction, magnitude, and significance.

```{r}

```

